Scalar Items in Embedded Position: {A}n Experimental Revisit

### 2012-07-07 text/paper/paper.tex
paper motivation section rewrite sec com

Summing up so far, three previous studies have presented different
pieces of evidence concerning putative local
readings. \citeauthor{ChemlaSpector2010:Experimental-Ev}'s results
suggest that local readings are not available, because they do not
affect a categorical truth-value judgments in a picture-verification
task. On the other hand,
\citeauthor{CliftonDube2010:Embedded-Implic}'s and
\citeauthor{ChemlaSpector2010:Experimental-Ev}'s design was suggested
to be sensitive enough to reveal effects of local readings: either by
a choice of the situation that better befits a given sentence
(\citeauthor{CliftonDube2010:Embedded-Implic}), or by a graded
truth-value judgment (\citeauthor{ChemlaSpector2010:Experimental-Ev}).

To account for all of these observations, it could be hypothesized
along with \citeauthor{CliftonDube2010:Embedded-Implic} and
\citeauthor{ChemlaSpector2010:Experimental-Ev} that local readings are
available, but so dispreferred that they do not affect simple
\emph{categorical} truth-value judgements. It could then furthermore
be argued, in line with traditionalism, that the reason why local
readings do not affect categorical judgements is because they are not
(strongly) truth-relevant speaker-intended meaning enrichments after
all, so that no case can be made for lexicalism or grammaticalism
\citep[c.f.][for arguments along these lines]{Tielvan-Tiel2012:Embedded-Scalar}.

Either position is cast into doubt by a pilot study of
ourselves.\dn{We should expand on the method in an appendix} We tried
to replicate \citeauthor{ChemlaSpector2010:Experimental-Ev}'s results,
using a simple categorial truth-value judgement task on pictures that
were exactly like those in Figures \ref{fig:AS-distinguishing-pics}
and \ref{fig:ES-distinguishing-pics}. Test sentences were German
cleft-constructions as in (\ref{bsp:AS-K1}) and
(\ref{bsp:ES-K1}).\fn{We concede that these cleft-constructions might
  unduly emphasize potential local readings, but that is not
  essential, because the main issue we are trying to address here is
  whether putative local readings \emph{could} affect categorical
  truth-value judgements.}
\begin{exe}
\ex \gll F\"{u}r \mymark{jedes} dieser Kinder gilt: es
  mag \mymark{einige} dieser Speisen. \label{bsp:AS-K1}\\
  For every {of these} children {it holds} it likes some {of
    these} dishes.\\
  For all of these children it holds that it likes some of these dishes.
\end{exe}
\begin{exe}
\ex \gll F\"{u}r \mymark{genau} \mymark{eines} dieser Kinder gilt: es
  mag \mymark{einige} dieser Speisen. \label{bsp:ES-K1}\\
  For exactly one {of these} children {it holds} it likes some {of
    these} dishes.\\
  For exactly one of these children it holds that it likes some of these dishes.
\end{exe}
Mean acceptance rates of 60 native German speakers for our target
sentences were:
\begin{center}
  \begin{tabular}{ccccccccc}
    \multicolumn{4}{c}{\as-sentences} &&
    \multicolumn{4}{c}{\es-sentences}     \\ \cmidrule(r){1-4} \cmidrule(r){6-9}
    false & literal & weak & strong && false & local & literal & all
    \\ \midrule 
    .9 & 72 & 79 & 90 && .8 & 20 & 71 & 93 
  \end{tabular}
\end{center}
Two observations are of interest here. Firstly, we observed 21\%
\emph{false}-judgements in the \as-weak condition, as well as 20\%
\emph{true}-judgements in the \es-local condition, contrary to the 0\%
in each case observed by
\citet{GeurtsPouscoulous2009:Embedded-Implic}. Secondly, the observed
difference between the \as-weak and the \as-strong conditions, on the
one had, and the \es-false and the \es-local condition, on the other
hand, both reached significance.\dn{spell out in appendix}

@@

These observed patterns suggest that local readings \emph{are}
available and \emph{can} have an effect even on categorical
truth-value judgements. However, this interpretation of the data is
problematic for reasons already addressed also by
\citet{ChemlaSpector2010:Experimental-Ev} and, more recently,
\citet{Tielvan-Tiel2012:Embedded-Scalar}. The problem is that the
present variants of truth-value judgment tasks all run risk of
conflating evidence about the availability or preference of readings
with properties of the pictorial material that are independent of the
reading to be assessed. This problem affects all of the studies
covered above, to some degree or other, because it follows almost
directly from the nature of the phenomenon under investigation and
therefore seems nearly impossible to avoid.

@@

@@

\end{figure}
%
Our pilot study revealed a difference in acceptability,
but a curious one. The sentence was judged \emph{true} in 90.3\% of
the cases for the situation in Figure~\ref{fig:global-EA} and in
98.3\% in situation \ref{fig:literal-EA}. The observed difference is
significant ($t = -3.8259$, $df = 322.375$, $p < 0.001$). This is a
curious finding, because, from the point of view of implicature
calculation, the only thing that could plausibly explain this
acceptability difference is the presence of a global implicature,
resulting from the negation of the corresponding
\es-sentence.\fn{Notice that (\ref{bsp:EA}) is the alternative to the
  corresponding \es-sentence needed for computing the putative global
  implicature. But these sentences are logically independent. So, it
  may be that the corresponding \es-sentence is an alternative also to
  sentences like (\ref{bsp:EA}). It may also be that being an
  alternative is not a symmetric notion. Be that as it may, our
  argument does not hinge on whether (\ref{bsp:EA}) does or does not
  actually have a global implicature of the alleged kind.} But such a
putative global implicature would be true in the
situation~\ref{fig:global-EA}, but false in the
situation~\ref{fig:literal-EA}. Consequently, the observed difference
in acceptability cannot be explained simply in terms of the hypothesis
of \citet{ChemlaSpector2010:Experimental-Ev} that acceptability is
proportional to the number of available true readings. Rather some
feature of the pictorial material that is independent of the putative
readings under scrutiny does affect truth-value judgements, even
categorical ones.

\citet{ChemlaSpector2010:Experimental-Ev} acknowledge the possibility
that the \emph{typicality} of a picture may affect graded truth-value
judgements. Stronger still, they show that graded judgments of
\as-sentences differ significantly for different instantiations of the
conditions in \ref{fig:AS-distinguishing-pics}, that vary in the
amount of connections without changing the assignment of truth values
to candidate readings. \citeauthor{ChemlaSpector2010:Experimental-Ev}
suggest that typicality of the pictorial material can account for
these differences, but submit that this does not explain away the
effect of the local reading. This is disputed by
\citet{Tielvan-Tiel2012:Embedded-Scalar}, who demonstrates that all of
\citeauthor{CliftonDube2010:Embedded-Implic}'s and
\citeauthor{ChemlaSpector2010:Experimental-Ev}'s data on \as-sentences
can be explained as typicality effects, dispensing the need to appeal
to any influence of local readings. We are sympathetic towards
\citeauthor{Tielvan-Tiel2012:Embedded-Scalar}'s innovative line of
reinterpretation of the data, but note, as
\citet{ChemlaSpector2010:Experimental-Ev} already observed, that his
typicality-based explanation does not extend to \es-sentences in an
obvious way\fn{For similar reasons, appeal to typicality cannot
  explain, as far as we can see, the preference for the situation in
  Figure~\ref{fig:literal-EA} over that in Figure\ref{fig:global-EA}
  given a sentence like (\ref{bsp:EA}).} and that it also does not
explain why we observed putative effects of typicality also in
categorical judgements in our pilot.

To conclude, the confounding effects of pictorial complexity and/or
typicality are real and affect even categorical judgements. The
available evidence about availability, preference and nature of local
readings is therefore still inconclusive. What is needed to complement
the picture is data from a design that elicits categorical judgements
that are as independent as possible of possibly conflating properties
of the pictorial material. Given the logical dependencies of the
relevant readings of \as- and \es-sentences this seems almost
impossible to achieve. As we will argue below we do, however, believe
it is.

On top of this, we would like to obtain more further information about
the relative preferences among all putative readings, so as to be able
to decide between theoretical positions, as outlined in
Section~\ref{sec:theories-predictions}. Neither of the previous
studies gives us that, either because they simply have not
investigated all the relevant comparisons
(\citet{GeurtsPouscoulous2009:Embedded-Implic},
\citet{CliftonDube2010:Embedded-Implic}), or because they cannot safely
derive the required cardinal information for danger of conflating
preferences for readings with preferences for pictures
(\citet{ChemlaSpector2010:Experimental-Ev}, our pilot study).

>>>>

Summing up so far, three previous studies have presented different
pieces of evidence concerning putative local
readings. \citeauthor{GeurtsPouscoulous2009:Embedded-Implic}'s results
suggest that local readings are not available, because they do not
affect a categorical truth-value judgments in a picture-verification
task. On the other hand,
\citeauthor{CliftonDube2010:Embedded-Implic}'s and
\citeauthor{ChemlaSpector2010:Experimental-Ev}'s design was suggested
to be sensitive enough to reveal effects of local readings: either by
a choice of the situation that better befits a given sentence
(\citeauthor{CliftonDube2010:Embedded-Implic}), or by a graded
truth-value judgment
(\citeauthor{ChemlaSpector2010:Experimental-Ev}).

The data made available by all of these three studies, if taken in
conjunction, can be interpreted in several ways. Here are the two most
obvious candidates that have been suggested in the literature.
Firstly, conventionalists or grammaticalists could maintain that the
latter two studies refute traditionalism because they show that local
readings are available, despite the fact that they do not affect
simple \mymark{categorical truth-value judgements}. This is
essentially the interpretation favored by
\citet{CliftonDube2010:Embedded-Implic}, as well as
\citet{ChemlaSpector2010:Experimental-Ev}. In contrast,
traditionalists could rebut this position by arguing that the very
reason why local readings do not affect categorical judgements is
because they are not truth-relevant speaker-intended meaning
enrichments after all. The traditionalist, who would like to argue in
this way, would then have to provide an alternative explanation for
the effects attributed to the availability of local readings in the
latter two studies. One such explanation, at least for the data
concerning \as-sentences, has recently been suggested by
\citet{Tielvan-Tiel2012:Embedded-Scalar}, who argues that the effects
of putative local readings in the latter mentioned studies are due to
\mymark{typicality effects}.\fn{Yet another possible interpretation
  could be given by literalism which admits to local readings but
  considers them dispreffered to literal readings. This position
  explains the tendentious effects reported in the latter two studies
  \emph{and} explains why local readings do not show in categorical
  judgements: literal readings are simply preferred, so that local
  readings only figure in tendentious choices, but not in categorical
  ones. We will come back to this position eventually.}

The main point we would like to make in this section is that neither
interpretation is fully satisfactory because the situation is even
more complicated than suggested so far. Therefore, further empirical
testing is required, but should be carefully designed so as to
circumnavigate, as far as possible, the delicate methodological
pitfalls the subject matter so unkindly provides.

In outline our position, which we will elaborate on in the following,
is this. On the one hand, we believe with
\citet{ChemlaSpector2010:Experimental-Ev} and
\citet{Tielvan-Tiel2012:Embedded-Scalar} that properties of the
pictorial material can affect subjects' judgements. We concede,
following \citet{Tielvan-Tiel2012:Embedded-Scalar}, that a notion of
typicality may partly be responsible. But we also submit, in line with
\citet{ChemlaSpector2010:Experimental-Ev}, that typicality might not
be all of the story. On the other hand, we present suggestive evidence
based on a pilot study of ourselves (to be addressed presently) that
even categorical judgements \emph{can} be affected by pictorial
properties. We conclude that further empirical evidence is needed and
that a follow-up study should meet the following requirements: (i) it
should be a categorical truth-value judgement task after all because
only that way the traditionalist claim that any observed effects are
not truth-relevant can be suitably addressed\dn{@Fabian: would a
  reference fit here to the book(?) that discusses why true/false
  judgements are best?} and (ii) it must make sure to minimize any
putative effects of the pictorial material as much as possible.

In support of this position, we would like to briefly discuss the
results of a small pilot study of ourselves.\fn{A word of
  motivation. We report the results of this pilot mainly for
  expository reasons. A better argument that categorical judgements
  can be indicative of local readings is given by our main
  experiment. But in order to motivate the readers that they should be
  interested in this experiment, we feel it is helpful to discuss the
  mildly suggestive, but ultimately inconclusive evidence of this
  pilot. Additionally, introducing the pilot also serves another
  purpose, namely the discussion of the results for sentences like
  (\ref{bsp:EA}). These help introduce in a tangible way how seemingly
  innocuous properties of the pictorial material can influence
  subjects' truth-value judgements, and why typicality might not be
  the whole explanation for these pictorial effects.} Our pilot tried
to replicate \citeauthor{GeurtsPouscoulous2009:Embedded-Implic}'s
results, using a simple categorical truth-value judgement task on
pictures that were exactly like those in Figures
\ref{fig:AS-distinguishing-pics} and
\ref{fig:ES-distinguishing-pics}.\dn{We should expand on the method in
  an appendix} Test sentences were German cleft-constructions as in
(\ref{bsp:AS-K1}) and (\ref{bsp:ES-K1}).\fn{We concede that these
  cleft-constructions might unduly emphasize potential local readings,
  perhaps because they induces a quotative reading. (We are grateful
  to Philippe Schlenker for raising this latter possibility.) But that
  is not essential, because the main issue we are trying to address
  here is whether putative local readings \emph{could} affect
  categorical truth-value judgements.}
\begin{exe}
\ex \gll F\"{u}r \mymark{jedes} dieser Kinder gilt: es
  mag \mymark{einige} dieser Speisen. \label{bsp:AS-K1}\\
  For every {of these} children {it holds} it likes some {of
    these} dishes.\\
  For all of these children it holds that it likes some of these dishes.
\end{exe}
\begin{exe}
\ex \gll F\"{u}r \mymark{genau} \mymark{eines} dieser Kinder gilt: es
  mag \mymark{einige} dieser Speisen. \label{bsp:ES-K1}\\
  For exactly one {of these} children {it holds} it likes some {of
    these} dishes.\\
  For exactly one of these children it holds that it likes some of these dishes.
\end{exe}
Mean acceptance rates of 60 native German speakers for our target
sentences were:
\begin{center}
  \begin{tabular}{cccc}
    \multicolumn{4}{c}{\as-sentences}     \\ \cmidrule(r){1-4}
    false & literal & weak & strong 
    \\ \midrule 
    .9 & 72 & 79 & 90 
  \end{tabular}
  \qquad
  \begin{tabular}{cccc}
    \multicolumn{4}{c}{\es-sentences}    \\ \cmidrule(r){1-4} 
    false & local & literal & all
    \\ \midrule 
    .8 & 20 & 71 & 93 
  \end{tabular}
\end{center}
% \begin{center}
%   \begin{tabular}{ccccccccc}
%     \multicolumn{4}{c}{\as-sentences} &&
%     \multicolumn{4}{c}{\es-sentences}     \\ \cmidrule(r){1-4} \cmidrule(r){6-9}
%     false & literal & weak & strong && false & local & literal & all
%     \\ \midrule 
%     .9 & 72 & 79 & 90 && .8 & 20 & 71 & 93 
%   \end{tabular}
% \end{center}
Two observations are of interest here. Firstly, we observed 21\%
\emph{false}-judgements in the \as-weak condition, as well as 20\%
\emph{true}-judgements in the \es-local condition, contrary to the 0\%
in each case observed by
\citet{GeurtsPouscoulous2009:Embedded-Implic}. Secondly, the observed
difference between the \as-weak and the \as-strong conditions, on the
one had, and the \es-false and the \es-local condition, on the other
hand, both reached significance.\dn{spell out in appendix} This
suggests that local readings \emph{are} available and \emph{can} have
an effect even on categorical truth-value judgements.

@@

However, the interpretation of data from judgement tasks is generally
problematic for reasons already addressed by
\citet{ChemlaSpector2010:Experimental-Ev} and, more recently,
\citet{Tielvan-Tiel2012:Embedded-Scalar}. The problem is that the
present variants of truth-value judgment tasks all run risk of
conflating evidence about the availability or preference of readings
with properties of the pictorial material that are independent of the
reading to be assessed. This problem affects all of the studies
covered above, to some degree or other, because it follows almost
directly from the nature of the phenomenon under investigation and
therefore seems nearly impossible to avoid.

@@

      \end{tikzpicture}
}

@@

\end{figure}
%
Our pilot study revealed a difference in acceptability, but a curious
one. The sentence was judged \emph{true} in 90.3\% of the cases for
the situation in Figure~\ref{fig:global-EA} and in 98.3\% in situation
\ref{fig:literal-EA}. The observed difference is significant ($t =
-3.8259$, $df = 322.375$, $p < 0.001$). This is a curious finding,
because, from the point of view of implicature calculation, the only
thing that could plausibly explain this acceptability difference is
the presence of a global implicature, resulting from the negation of
the corresponding \es-sentence.\fn{Notice that (\ref{bsp:EA}) is the
  alternative to the corresponding \es-sentence needed for computing
  the putative global implicature of that \es-sentence. But since
  these sentences are logically independent, it may be assumed that
  the corresponding \es-sentence is an alternative also to sentences
  like (\ref{bsp:EA}). It may also be that \emph{being-an-alternative}
  is not a symmetric notion. Be that as it may, our argument does not
  hinge on whether (\ref{bsp:EA}) does or does not actually have a
  global implicature of the alleged kind.} But such a putative global
implicature would be true in the situation~\ref{fig:global-EA}, but
false in the situation~\ref{fig:literal-EA}. Consequently, the
observed difference in acceptability cannot be explained simply in
terms of the hypothesis of \citet{ChemlaSpector2010:Experimental-Ev}
that acceptability is proportional to the number of available true
readings. Rather some feature of the pictorial material that is
independent of the putative readings under scrutiny does affect
truth-value judgements, even categorical ones.

\citet{ChemlaSpector2010:Experimental-Ev} acknowledge the possibility
that the \emph{typicality} of a picture may affect graded truth-value
judgements. Stronger still, they show that graded judgments of
\as-sentences differ significantly for different instantiations of the
conditions in \ref{fig:AS-distinguishing-pics}, that vary in the
amount of connections without changing the assignment of truth values
to candidate readings. \citeauthor{ChemlaSpector2010:Experimental-Ev}
suggest that typicality of the pictorial material can account for
these differences, but submit that this does not explain away the
effect of the local reading. This is disputed by
\citet{Tielvan-Tiel2012:Embedded-Scalar}, who demonstrates that all of
\citeauthor{CliftonDube2010:Embedded-Implic}'s and
\citeauthor{ChemlaSpector2010:Experimental-Ev}'s data on \as-sentences
can be explained as typicality effects, dispensing the need to appeal
to any influence of local readings. We are sympathetic towards
\citeauthor{Tielvan-Tiel2012:Embedded-Scalar}'s innovative line of
reinterpretation of the data, but note, as
\citet{ChemlaSpector2010:Experimental-Ev} already observed, that his
typicality-based explanation does not extend to \es-sentences in an
obvious way\fn{For similar reasons, appeal to typicality cannot
  explain, as far as we can see, the preference for the situation in
  Figure~\ref{fig:literal-EA} over that in Figure~\ref{fig:global-EA}
  given a sentence like (\ref{bsp:EA}).} and that it also does not
explain why we observed putative effects of typicality also in
categorical truth-value judgements in our pilot.

To conclude, the confounding effects of pictorial complexity and/or
typicality are real and affect even categorical truth-value
judgements. The available evidence about availability, preference and
nature of local readings is therefore still inconclusive. What is
needed to complement the picture is data from a design that elicits
categorical truth-value judgements that are as independent as possible
of possibly conflating properties of the pictorial material. Given the
logical dependencies of the relevant readings of \as- and
\es-sentences this seems almost impossible to achieve. As we will
argue below we do, however, believe it is.

On top of this, we would like to obtain further information about
the relative preferences among all putative readings, so as to be able
to decide between theoretical positions, as outlined in
Section~\ref{sec:theories-predictions}. Neither of the previous
studies gives us that, either because they simply have not
investigated all the relevant comparisons
(\citet{GeurtsPouscoulous2009:Embedded-Implic},
\citet{CliftonDube2010:Embedded-Implic}), or because they cannot safely
derive the required cardinal information for danger of conflating
preferences for readings with preferences for pictures
(\citet{ChemlaSpector2010:Experimental-Ev}, our pilot study).
